---
<!DOCTYPE html>
<html>

<body>

<head>
<style>
  p.solid {border-style: solid;}
  <title>Getting Data Into Browser using R </title>
</style>
</head>

 
<div id="header"  class="solid" >
<h2><center> Getting Data Into Browser using R   </center> </h1>
</div>

</div>

<div id="section1"  <center>>
<p >
Output format: HTML.  
Data Source : NYPD vehicle collision data.    
Data Transformation: Year 2014: Summary of persons injured by borough.    
Chart: Column chart comparing injuries by borough.   
</p>


```{r setup, echo = T, cache = F, message = F}
require(knitr)
nypdvcrashSummary <-read.csv("c:/Aarti/Columbia Data Sciences/EDAV/RProject/data/NYPDSummary_Byborough.csv",header=T,sep=",")
require("ggplot2")

#BAR PLOTS
library(MASS)
library(plyr)

(bp1 <- ggplot(data=nypdvcrashSummary, 
        aes(x=BOROUGH,y=Frequency,
        fill=factor(Type))) 
     +  geom_bar(stat="identity", position=position_dodge()) 
     +  ggtitle("summary - Persons Injured By Boroughs")
     +  geom_text(aes(label=Frequency), vjust=-1, size=3)
     +  theme(plot.title=element_text(size=rel(1), face="bold"))
     +  theme(axis.title=element_text(size=15)))
```
</center>
</div>

<div id="sectionw" <center>>
<p>
Data Source : NYPD vehicle collision data.   
Data Transformation: Year 2014: Summary of persons injured by borough and vehicle type in accident.      
Chart: Column chart comparing injuries by borough by vehicle types.
</p>
</p>


``` {r echo = T , cache = F, message = F}


nypdvcrashSummary2 <-read.csv("c:/Aarti/Columbia Data Sciences/EDAV/RProject/data/NYPDSummary_ByVehicleType.csv",header=T,sep=",")
require("ggplot2")

#BAR PLOTS
library(MASS)
library(plyr)

(bp1 <- ggplot(data=nypdvcrashSummary2, 
               aes(x=BOROUGH,y=Frequency,fill=factor(Type))) + 
               geom_bar(stat="identity", position=position_dodge()) + 
   ggtitle("Person injuries in 2014 by Vehicle Type ")
   + geom_text(aes(label=Frequency), vjust=0, size=3)
   + theme(plot.title=element_text(size=rel(1), face="bold"))
   + theme(axis.title=element_text(size=15)))



```

</center>
</div>

<div id="footer">
Footer:  Copyright columbia.edu

</div>
</body>
</html>
